---
title: "Temporary Preview: Meta-Analysis Plot Redesigns"
format:
  html:
    toc: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
library(metafor)
library(ggplot2)

set.seed(20260419)

bg <- "#FFFFFF"

simulate_smd_meta <- function(k = 12, n_min = 30, n_max = 100, m1 = 0.4, m2 = 0, sd1 = 1, sd2 = 1, method = "REML") {
  dat <- data.frame(yi = numeric(0), vi = numeric(0), n = integer(0), study = character(0))
  for (i in seq_len(k)) {
    n_i <- sample(n_min:n_max, 1)
    x <- rnorm(n_i, mean = m1, sd = sd1)
    y <- rnorm(n_i, mean = m2, sd = sd2)
    es <- metafor::escalc(
      n1i = n_i, n2i = n_i,
      m1i = mean(x), m2i = mean(y),
      sd1i = sd(x), sd2i = sd(y),
      measure = "SMD"
    )
    dat[i, "yi"] <- es$yi
    dat[i, "vi"] <- es$vi
    dat[i, "n"] <- n_i
    dat[i, "study"] <- paste("Study", i)
  }
  res <- metafor::rma(yi, vi, data = dat, method = method)
  list(dat = dat, res = res)
}

simulate_or_meta <- function(k = 12, n_min = 30, n_max = 80, pr1 = 0.70, pr2 = 0.50) {
  ai <- numeric(k)
  bi <- numeric(k)
  ci <- numeric(k)
  di <- numeric(k)
  n_each <- integer(k)

  for (i in seq_len(k)) {
    n_i <- sample(n_min:n_max, 1)
    x <- rbinom(n_i, 1, pr1)
    y <- rbinom(n_i, 1, pr2)
    ai[i] <- sum(x == 1)
    bi[i] <- sum(x == 0)
    ci[i] <- sum(y == 1)
    di[i] <- sum(y == 0)
    n_each[i] <- n_i
  }

  dat <- data.frame(ai = ai, bi = bi, ci = ci, di = di, n = n_each, study = paste("Study", seq_len(k)))
  es <- metafor::escalc(measure = "OR", ai = ai, bi = bi, ci = ci, di = di, data = dat)
  dat$yi <- es$yi
  dat$vi <- es$vi
  res <- metafor::rma(yi, vi, data = dat)
  list(dat = dat, res = res)
}

make_heterogeneity_example <- function(k = 30) {
  dat <- data.frame(yi = numeric(0), vi = numeric(0), subgroup = character(0), study = character(0))

  for (i in seq_len(k / 2)) {
    n_i <- sample(30:100, 1)
    x <- rnorm(n_i, 0.5, 1)
    y <- rnorm(n_i, 0.0, 1)
    es <- metafor::escalc(n1i = n_i, n2i = n_i, m1i = mean(x), m2i = mean(y), sd1i = sd(x), sd2i = sd(y), measure = "SMD")
    dat[i, 1:2] <- es
    dat[i, "subgroup"] <- "Manipulation A"
    dat[i, "study"] <- paste("Study", i)
  }

  for (i in (k / 2 + 1):k) {
    n_i <- sample(30:100, 1)
    x <- rnorm(n_i, 0.2, 1)
    y <- rnorm(n_i, 0.0, 1)
    es <- metafor::escalc(n1i = n_i, n2i = n_i, m1i = mean(x), m2i = mean(y), sd1i = sd(x), sd2i = sd(y), measure = "SMD")
    dat[i, 1:2] <- es
    dat[i, "subgroup"] <- "Manipulation B"
    dat[i, "study"] <- paste("Study", i)
  }

  dat <- dat[sample(nrow(dat)), ]
  rownames(dat) <- NULL
  res <- metafor::rma(yi, vi, data = dat)
  list(dat = dat, res = res)
}

forest_df_from_rma <- function(res, dat, level = 0.95, label_col = "study") {
  z <- qnorm(1 - (1 - level) / 2)
  out <- data.frame(
    study = dat[[label_col]],
    yi = dat$yi,
    vi = dat$vi,
    sei = sqrt(dat$vi)
  )
  out$ci_lb <- out$yi - z * out$sei
  out$ci_ub <- out$yi + z * out$sei
  out$weight <- 1 / out$vi
  out$weight_scaled <- out$weight / max(out$weight)
  out$order <- rev(seq_len(nrow(out)))

  pooled <- data.frame(
    study = "Pooled",
    yi = as.numeric(res$b[1]),
    ci_lb = as.numeric(res$ci.lb),
    ci_ub = as.numeric(res$ci.ub),
    order = 0
  )

  list(studies = out, pooled = pooled)
}

plot_journal_ggforest <- function(res, dat, xlab = "Hedges g", title = "Custom ggplot forest") {
  fd <- forest_df_from_rma(res, dat)
  p <- ggplot(fd$studies, aes(y = order, x = yi)) +
    geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "#777777") +
    geom_segment(aes(x = ci_lb, xend = ci_ub, yend = order), linewidth = 0.55, color = "#303030") +
    geom_point(aes(size = weight_scaled), shape = 15, color = "#111111") +
    geom_segment(data = fd$pooled, aes(x = ci_lb, xend = ci_ub, y = order, yend = order), linewidth = 1.2, color = "#1B4D3E") +
    geom_point(data = fd$pooled, aes(x = yi, y = order), color = "#1B4D3E", shape = 18, size = 3.5, inherit.aes = FALSE) +
    scale_y_continuous(
      breaks = c(fd$studies$order, 0),
      labels = c(fd$studies$study, "Pooled"),
      expand = expansion(mult = c(0.02, 0.06))
    ) +
    scale_size_continuous(range = c(1.8, 5.5), guide = "none") +
    labs(x = xlab, y = NULL, title = title) +
    coord_cartesian(clip = "off") +
    theme_bw(base_size = 11) +
    theme(
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold"),
      axis.text.y = element_text(size = 9)
    )

  p
}

old_data <- simulate_smd_meta(k = 12, method = "FE")
old_data_small <- simulate_smd_meta(k = 4, n_min = 30, n_max = 80, m1 = 0.4, method = "FE")
meta_bias_dat <- data.frame(
  yi = c(0.52, 0.45, 0.22, 0.31, 0.53),
  vi = c(0.038, 0.0399, 0.031, 0.034, 0.031),
  study = c("A", "B", "C", "D", "E")
)
meta_bias_all <- metafor::rma(yi, vi, data = meta_bias_dat, slab = meta_bias_dat$study)
meta_bias_sig <- metafor::rma(yi, vi, data = meta_bias_dat[c(1, 2, 5), ], slab = c("A", "B", "E"))
or_example <- simulate_or_meta()
het_example <- make_heterogeneity_example()
```

## Goal

This temporary file does two things:

1. Reproduces the current style of forest plots used in the book ("old").
2. Shows redesign options that are clearer and journal-oriented with minimal dependencies.

Package policy used here:

- No new required dependency for redesigns (uses `metafor` + `ggplot2`, already used in the book).
- One optional extra package explored: `meta`.

## A. Current Book-Style Forest Plots (Old)

### A1. Simple 4-study forest (similar to Chapter 7)

```{r old-simple-4, fig.height=4.8, fig.width=8}
par(mar = c(4, 1, 3, 1), bg = bg)
metafor::forest(old_data_small$res, top = 0)
title("Old style: simple forest")
```

### A2. Publication-bias example with two stacked forests (similar to Chapter 6)

```{r old-metabias, fig.height=8, fig.width=8}
par(mfrow = c(2, 1), bg = bg)
par(mar = c(4, 4, 2, 1))
metafor::forest(
  meta_bias_all,
  xlab = "Cohen's d",
  efac = c(0, 4),
  lty = c(1, 1, 0),
  cex = 0.8,
  shade = TRUE,
  rowadj = 0,
  psize = 1,
  colout = "#333333",
  xlim = c(-0.4, 1.5),
  alim = c(-0.2, 1),
  textpos = c(-0.4, 1.5),
  header = TRUE
)
metafor::forest(
  meta_bias_sig,
  xlab = "Cohen's d",
  efac = c(0, 4),
  lty = c(1, 1, 0),
  cex = 0.8,
  shade = TRUE,
  rowadj = 0,
  psize = 1,
  colout = "#333333",
  xlim = c(-0.4, 1.5),
  alim = c(-0.2, 1),
  textpos = c(-0.4, 1.5),
  header = TRUE
)
par(mfrow = c(1, 1))
```

### A3. 12-study simulated forest (similar to Chapter 11)

```{r old-12, fig.height=6.5, fig.width=8}
par(mar = c(4, 4, 2, 2), bg = bg)
metafor::forest(old_data$res)
title("Old style: 12-study forest")
```

### A4. Heterogeneity forest with reference lines (similar to Chapter 11)

```{r old-heterogeneity, fig.height=8.5, fig.width=8}
par(mar = c(4, 1, 2, 1), bg = bg)
metafor::forest(
  het_example$res,
  efac = c(0, 3),
  lty = c(1, 1, 0),
  cex = 0.8,
  shade = TRUE,
  rowadj = 0,
  psize = 1,
  colout = "#333333",
  xlim = c(-2, 2.5),
  alim = c(-1, 1.5),
  textpos = c(-2, 2.5),
  at = c(-1, -0.5, 0, 0.5, 1, 1.5),
  refline = c(0, 0.2, 0.5),
  header = TRUE
)
title("Old style: heterogeneity forest")
```

## B. Redesign Option 1: Better `metafor::forest()` Styling (No New Dependency)

These are still base `metafor` plots, but tuned for cleaner journal-like output.

```{r redesign-metafor-clean, fig.height=6.8, fig.width=9}
par(mar = c(4, 6, 2, 2), bg = "white")
metafor::forest(
  old_data$res,
  slab = old_data$dat$study,
  xlab = "Standardized mean difference (Hedges g)",
  cex = 0.9,
  psize = 1.2,
  colout = "#1a1a1a",
  col = "#1a1a1a",
  lwd = 1,
  shade = FALSE,
  refline = 0,
  header = c("Study", "Effect [95% CI]"),
  digits = 2,
  xlim = c(-1.5, 2.2),
  alim = c(-1.2, 1.8),
  at = seq(-1.0, 1.5, 0.5)
)
title("Redesign 1: Clean metafor style")
```

```{r redesign-metafor-or, fig.height=6.8, fig.width=9}
par(mar = c(4, 6, 2, 2), bg = "white")
metafor::forest(
  or_example$res,
  slab = or_example$dat$study,
  ilab = cbind(or_example$dat$ai, or_example$dat$bi, or_example$dat$ci, or_example$dat$di),
  ilab.xpos = c(-8.2, -7.3, -6.4, -5.5),
  xlim = c(-9.2, 4.8),
  alim = c(-2.0, 2.0),
  at = seq(-2, 2, 1),
  xlab = "log(odds ratio)",
  psize = 1.1,
  cex = 0.85,
  refline = 0,
  header = c("Study", "log(OR) [95% CI]"),
  digits = 2
)
text(c(-8.2, -7.3, -6.4, -5.5), length(or_example$dat$study) + 2.1, c("E+", "E-", "C+", "C-"), cex = 0.85, font = 2)
title("Redesign 1b: Clean OR forest with count columns")
```

## C. Redesign Option 2: Custom `ggplot2` Forest (No New Dependency)

This option gives full styling control and consistent typography with other ggplot figures.

```{r redesign-ggplot, fig.height=6.5, fig.width=9}
plot_journal_ggforest(old_data$res, old_data$dat, xlab = "Hedges g", title = "Redesign 2: Custom ggplot forest")
```

```{r redesign-ggplot-heterogeneity, fig.height=8, fig.width=9}
p <- plot_journal_ggforest(het_example$res, het_example$dat, xlab = "Hedges g", title = "Redesign 2b: Heterogeneity-focused ggplot forest")
p +
  geom_vline(xintercept = 0.2, linetype = "dotted", color = "#7A0019", linewidth = 0.6) +
  geom_vline(xintercept = 0.5, linetype = "dotted", color = "#7A0019", linewidth = 0.6)
```

## D. Redesign Option 3: `meta` Package Forest (One Extra Dependency, Optional)

This explores a dedicated forest API with compact publication defaults.

```{r redesign-meta-package, fig.height=6.5, fig.width=9}
if (requireNamespace("meta", quietly = TRUE)) {
  m_obj <- meta::metagen(
    TE = old_data$dat$yi,
    seTE = sqrt(old_data$dat$vi),
    studlab = old_data$dat$study,
    sm = "SMD"
  )

  meta::forest(
    m_obj,
    xlab = "Standardized mean difference",
    leftcols = c("studlab"),
    leftlabs = c("Study"),
    rightcols = c("effect", "ci"),
    rightlabs = c("Effect", "95% CI"),
    col.square = "#111111",
    col.diamond = "#1B4D3E",
    col.diamond.lines = "#1B4D3E",
    col.study = "#111111",
    colgap.forest.left = "8mm",
    colgap.forest.right = "8mm"
  )
} else {
  cat("Package 'meta' is not installed. Install with install.packages('meta') to view this option.\n")
}
```

## Recommendation Summary

Based on readability, flexibility, and dependency constraints:

1. If you want minimal change and no new dependencies: use **Option 1** (cleaned `metafor::forest`).
2. If you want the highest visual quality and consistency with your ggplot figures: use **Option 2** (custom `ggplot2` forest template).
3. If you are open to one extra dependency for forest-plot ergonomics: try **Option 3** (`meta` package) and compare.

## Suggested Next Step

Once you pick an option, we can create one reusable plotting helper in [R/](R/) and replace forest-plot chunks chapter by chapter.
